---
title: "Introduction to the `cpass` package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to the `cpass` package}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = FALSE
)
```


# Installation of the `cpass` package

Currently, the `cpass` package is hosted on the package author's github repository.

To install it from github, install and load the package `devtool` by running:

```{r devtools, eval = FALSE}
install.packages("devtools")
```


```{r loading-devtools, warning=FALSE, eval = FALSE}
library(devtools)
```


then install and load the `cpass` package by running:

```{r cpass-package, eval = FALSE}
devtools::install_github("lasy/cpass", dependencies = TRUE)
```

```{r loading-cpass-package}
library(cpass)
```


Note that the `cpass` package depends on several other packages (mostly packages from the [`tidyverse` suite](https://www.tidyverse.org/)).

This vignette also uses `tidyverse` functions so it is recommended to install and load the `tidyverse` suite by running:



```{r tidyverse, eval = FALSE}
install.packages("tidyverse")
```



```{r loading-tidyverse, warning=FALSE, eval  = FALSE}
library(tidyverse)
```

Alternatively, individual packages can be loaded independently:

```{r loading-pcks, warning=FALSE}
library(dplyr)
library(ggplot2)
library(purrr)
library(tibble)
library(tidyr)
library(stringr)
library(magrittr)
```


# Demo dataset

To explore the `cpass` package, we will use a demo dataset with the data (daily ratings) from 20 subjects. This demo dataset is attached to the `cpass` package and documented in `?PMDD_data`. 

To load this data into the environment, run:

```{r demo-dataset}
d = PMDD_data

dim(d)
head(d)
```


# Overview of the package functions 

The `cpass` package contains 3 types of functions:

1. a data formatting function `as_cpass_data()` which first checks the format of the data you would like to analyse with the `cpass` package, then formats it for diagnosis and/or visualizations

2. a PMDD/MRMD/PME diagnosis function `cpass()`

3. visualization functions: `plot_subject_obs()`, `plot_subject_cycle_obs()`, `plot_subject_dx()` and `plot_subject_data_and_dx()`

A typical `cpass` workflow consists of

* loading (and formatting) raw data from a `.csv` or `.xls` file

* preparing them for diagnosis with the `as_cpass_data()`

* running the `cpass()` function on these data and extract some general statistics about the diagnoses

* visually explore the data of specific subjects with the visualization functions.


# Package demo

## Formating raw data into `cpass_data`

The function `as_cpass_data()` transforms raw data into `cpass_data`. The data provided as argument to `as_cpass_data()` must be in long format, _i.e._ one row per subject, cycle, day and DRSP (daily record of severity of problems) item (*i.e.*, the symptom diary used to collect the data for a C-PASS diagnosis). 

```{r as-cpass-data}
input = as_cpass_data(d, sep_event = "menses")
```


For the full documentation, type `?as_cpass_data` in the `R console`.


## Running the CPASS procedure for PMDD/MRMD/PME diagnosis

Now that the data is formatted, it can be passed to the `CPASS()` function. This function will run the CPASS procedure for PMDD/MRMD (and optionally PME) diagnoses:

```{r CPASS}
diagnosis = cpass(input)
```

For the full documentation, type `?CPASS` in the `R console`.

The `CPASS` function returns a list of 6 tables:

```{r CPASS-results}
names(diagnosis)
```


Let's have a look at the first table:

```{r}
diagnosis$subject_level_diagnosis
```


This table has the diagnoses of the 20 subjects.

```{r}
table(diagnosis$subject_level_diagnosis$dx)
```


3 subjects (~15%) have a PMDD diagnosis, 3 others have a MRMD diagnosis and 9 subjects (~45%) have no diagnosis based on their reported symptoms.

The remaining 5 subjects did not have sufficient data to make a diagnosis:

```{r}

diagnosis$subject_level_diagnosis %>% 
  filter(is.na(dx))

```

We see that each of these 5 subjects only reported DRSP for one cycle (instead of the two required).


Now, we can dive deeper into the data of particular subjects.
For example, we can check the cycle-level diagnosis of subject `#47` who reported DRSP scores for 3 cycles.

```{r}
diagnosis$cycle_level_diagnosis %>% filter(subject == 47)
```

The second cycle of this subject fulfilled the criteria for a cycle-level PMDD diagnosis. The symptoms reported in the other cycles did not meet the criteria for PMDD nor for MRMD.

We can dive even deeper and look at the DRSP items during this cycle #2:

```{r}
diagnosis$item_level_diagnosis %>% filter(subject == 47, cycle == 2)
```

As this is a large table which provides the criteria details for each item, we may only consider the following columns for now:

```{r}

diagnosis$item_level_diagnosis %>% 
  filter(subject == 47, cycle == 2) %>% 
  select(subject, cycle, item, item_meets_PMDD_criteria, DSM5_SYMPTOM_DOMAIN) %>%  
  as.data.frame()

```

And we observe that this subject mostly experienced severe anxiety, inability to control their mood, anger and feeling of overwhelm in the premenstrual phase of cycle 2.


## Visualizing subjects data

While the subject data and diagnoses can be explored by inspecting these tables, the `cpass` package comes with visualization functions which may help in the investigation of a specific subject's data.

### Visualizing subject's diagnoses

For example, one can have an overview of a subject diagnosis with the `plot_subject_dx()` function.

For the same subject #47, we run:

```{r plot-subject-diagnosis, fig.width=14, fig.height=10, out.width='100%'}

plot_subject_dx(input %>% dplyr::filter(subject == 47))

```

The top panel provides the subject ID, their diagnosis, the number of cycles for which some scores were provided and the number of cycles that had enough data to be included in the diagnosis procedure. It also provides the number of cycles with a PMDD or a MRMD diagnosis.

### Visualizing subject's reported scores

From the `cycle_level_dx` table from this subject, we know that cycle #2 fulfilled the PMDD criteria while other cycles did not.

To explore the scores reported during the 2nd cycle, the function `plot_subject_cycle_obs()` can be used.
This function visualizes the reported score for a given cycle:


```{r plot-cycle, fig.height=12, fig.width=14, out.width='100%'}
plot_subject_cycle_obs(data = input %>% filter(subject == 47, cycle == 2))
```


The two left panels visualize the reported scores with the x-axis being the cycle-days (time) and each line of the y-axis representing one DRSP item. The score of each item for each day is written on the corresponding (x,y) rectangle. The color of the rectangle is proportional to the score. Low scores (1) are gray, high scores (6) are red. Missing scores are represented by transparent rectangles. The panels on the right visualize the items or domains or cycle that fulfill the PMDD criteria. If they do not meet the criteria, their corresponding rectangle is light-gray. If they meet the criteria, they are red. Items that are excluded from the diagnosis procedure are in dark-gray.

For this subject, we can compare their second cycle (above) with their third cycle:

```{r plot-cycle-3, fig.height=12, fig.width=14, out.width='100%'}
plot_subject_cycle_obs(data = input %>% filter(subject == 47, cycle == 3))
```

Note that the reported scores can be visualized without the diagnosis with the option `add_diagnosis = FALSE`

```{r plot-cycle-no-d, fig.height=12, fig.width=10, out.width='100%'}
plot_subject_cycle_obs(data = input %>% filter(subject == 47, cycle == 2), add_diagnosis = FALSE)
```

And that the color of the high score can be changed with the option `color_max_score = "desired_color"`:


```{r plot-cycle-color, fig.height=12, fig.width=14, out.width='100%'}
plot_subject_cycle_obs(data = input %>% filter(subject == 47, cycle == 2), color_max_score = "turquoise")
```

It is also possible to visualize all cycles of a user at once with the function `plot_subject_obs()`

```{r plot-obs, fig.height=36, fig.width=14, out.width='100%'}
plot_subject_obs(data = input %>% filter(subject == 47))
```

### Visualizing subject's reported scores and diagnosis

And finally, the function `plot_subject_data_and_dx()` is useful to visualize at once all reported scores and the diagnoses summaries of a given subject.

Because this function may return large visualizations, by default, it prints the visualization to a pdf that has an appropriate size (*i.e.*, a height proportional to the number of cycles reported by the subject).

By default, the pdf is saved in the current directory and the filename specifies the subject ID but both the path and the pdf name can be specified via the options `pdf_path` and `pdf_name`.

The pdf option can also be turned off with the option `save_as_pdf = FALSE`. In this case, calling the function prints the visualization in R.

If the function is called with the option `save_as_pdf = TRUE` (default), the visualization is returned silently but could be caught by a variable (e.g. `viz = plot_subject_data_and_dx(data); viz`).

```{r plot-all, fig.height=50, fig.width=14, out.width='100%'}

plot_subject_data_and_dx(data = input %>% filter(subject == 47), save_as_pdf = FALSE)

```


