---
title: "Introduction to the 'cpass' package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{cpass_vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


# Installation of the `cpass` package

Currently, the `cpass` package is hosted on the package author github repository.

To install it from this github repository, install and load the package `devtool` by running:

```{r devtools, eval = FALSE}
install.packages("devtools")
```

```{r loading-devtools}
library(devtools)
```


then install and load the `cpass` package by running:

```{r cpass-package, eval = FALSE}
devtools::install_github("lasy/cpass", dependencies = TRUE)
```

```{r loading-cpass-package}
library(cpass)
```


Note that the `cpass` package depends on several other packages (mostly packages from the [`tidyverse` suite](https://www.tidyverse.org/)).

This vignette also uses many of the `tidyverse` functions so it is recommended to install and load the `tidyverse` suite by running:

```{r tidyverse, eval = FALSE}
install.packages("tidyverse")
```

```{r loading-tidyverse}
library(tidyverse)
```


# Demo dataset

To explore the `cpass` package, we will use a demo dataset with the data (daily ratings) from 20 subjects. This demo dataset is attached to the `cpass` package and documented in `?PMDD_data`. 

To load this data into the environment, run

```{r demo-dataset}
d = PMDD_data

dim(d)
head(d)
```


# Overview of the package functions 

The `cpass` package contains 3 types of functions:

1. a data formatting function `as_cpass_data()` which first checks the format of the data you would like to analyse with the `cpass` package, then formats it for diagnosis and/or visualizations

2. a PMDD/MRMD/PME diagnosis function `CPASS()`

3. visualization functions: `plot_subject_obs()`, `plot_subject_cycle_obs()`, `plot_subject_diagnosis()` and `plot_subject_data_and_diagnosis()`

A typical `cpass` workflow consists in

* loading (and formatting) raw data from a `.csv` or `.xls` file

* preparing them for diagnosis with the `as_cpass_data()`

* running the `CPASS()` function on these data and extract some general statistics about the diagnoses

* visualy explore the data of specific subjects with the visualization functions.


# Package demo

## Formating raw data into cpass_data

The function `as_cpass_data()` transforms raw data into `cpass_data`. The data provided as argument to `as_cpass_data()` must be in long format, _i.e._ one row per subject, cycle, day and DRSP item. 

```{r as-cpass-data}

input = as_cpass_data(d)

```


For the full documentation, type `?as_cpass_data` in the `R console`.


## Running the CPASS procedure for PMDD/MRMD/PME diagnosis

Now that the data is formatted, it can be passed to the `CPASS()` function, which will run the CPASS procedure for PMDD/MRMD (and optionally PME) diagnoses:

```{r CPASS}

diagnosis = CPASS(input)

```

For the full documentation, type `?CPASS` in the `R console`.

The `CPASS` function returns a list of 6 tables:

```{r CPASS-results}

names(diagnosis)

```


Let's have a look at the first table:

```{r}
diagnosis$SUBJECT_level_diagnosis
```


This table has the overall diagnoses for the 20 subjects.

```{r}
table(diagnosis$SUBJECT_level_diagnosis$dx)
```


3 subjects (~15%) have a PMDD diagnosis, the same number a MRMD diagnosis and 9 subjects (~45%) have no diagnosis for their reported symptoms.

We can dive deeper into the data of particular subjects.
For example, we can check the cycle-level diagnosis of subject `#47` for which there was 3 cycles of data.

```{r}
diagnosis$CYCLE_level_diagnosis %>% filter(SUBJECT == 47)
```

Once cycle of this subject (cycle #2) received a PMDD diagnosis while the symptoms reported in the other cycles did not meet the criteria for PMDD or MRMD.

We can dive even deeper and look at the DRSP items during this cycle #2:

```{r}
diagnosis$DRSP_level_diagnosis %>% filter(SUBJECT == 47, CYCLE == 2)
```

As this is a large table which provides the details of the criteria met for each item, we may only consider the following columns for now:

```{r}
diagnosis$DRSP_level_diagnosis %>% filter(SUBJECT == 47, CYCLE == 2) %>% select(SUBJECT, CYCLE,DRSP, DRSP_meets_criteria, DSM5_SYMPTOM_DOMAIN)
```

And we observe that this subject mostly experienced severe anxiety, inability to control their mood, anger and feeling of overwhelm in the pre-menstrual phase of cycle 2.


## Visualizing subjects data

While the subject diagnoses can all be explored via the tables in the `diagnosis` variable, the `cpass` package comes with visualization functions which may help in the exploration of specific subject's data.

For example, one can have an overview of a subject diagnosis with the `plot_subject_diagnosis()` function.

For the same subject #47, we run:

```{r plot-subject-diagnosis, fig.width=10, fig.height=8}
plot_subject_diagnosis(input %>% filter(SUBJECT == 47))
```




